---
title: "The Role of Race and Mental Illness Diagnosis on Stigmatization of Homeless Individuals"
shorttitle: "STIGMA, HOMELESSNESS, MENTAL ILLNESS, AND RACE"
author:
  - name: Karen Veronica Becerra
    corresponding: true
    orcid: 0009-0006-4967-0955
    email: kvbecerra@uchicago.edu
    affiliations:
      - id: uchicago
        name: "The University of Chicago"
        group: MAPSS Psychology
        department: Department of Psychology
        address: 5848 S. University Avenue
        city: Chicago
        region: IL
        country: USA
        postal-code: 60637
blank-lines-above-author-note: 2
author-note:
  status-changes: ~
  disclosures:
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
abstract: "Homelessness in the United States is a persistent problem that can have serious implications on the well-being of homeless individuals. The present study focused on the role of race and mental illness diagnosis on the stigmatization of homeless individuals, specifically looking at the outcomes of the Attribution Questionnaire. This questionnaire assessed the aspects of social distance, blame, dangerousness, concern, and willingness to help of 215 participants varying in ages across adulthood. The study was a self-paced online form that used six experimental vignettes. The results indicated that there were no significant interactions of race x diagnosis on stigmatization. Additionally, race had no significant main effects, suggesting it was not a significant factor for stigmatization of homeless individuals. However, there were some significant main effects of diagnosis. Findings might suggest that future work in reducing mental illness stigma and increasing education could help decrease stigmatization of the homeless population."
keywords: [Homelessness, Stigmatization, Race, Mental Illness, Diagnosis, Attribution Questionnaire]
impact-statement: ~
floatsintext: true
numbered-lines: false
bibliography: "Stigmatization.bib"
suppress-title-page: false
link-citations: true
mask: false
masked-citations: []
draft-date: false
lang: en
language:
  citation-last-author-separator: "and"
  citation-masked-author: "Masked Citation"
  citation-masked-date: "n.d."
  citation-masked-title: "Masked Title"
  email: "Email"
  title-block-author-note: "Author Note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: ~
  title-impact-statement: "Impact Statement"
  references-meta-analysis: ~
format:
  apaquarto-docx: 
    toc: false
  apaquarto-html: 
    toc: true
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
fontsize: 12pt
mainfont: "Times New Roman"
donotrepeattitle: true
execute: 
  freeze: auto
  working-directory: project
---

```{r}
#| label: setup
#| include: false
library(conflicted)
library(tidyverse)
library(flextable)
library(ftExtra)
library(knitr)
conflicts_prefer(dplyr::filter, .quiet = TRUE)
conflicts_prefer(flextable::separate_header, .quiet = TRUE)

```
Homelessness in the United States and the struggle to give individuals adequate housing is a persistent problem. Before the Covid -19 pandemic, the number of homeless individuals was on the rise with 568,000 individuals experiencing homelessness in 2019, an increase of 15,000 from the previous year[@frostHomelessnessWasRise2020]^[@frostHomelessnessWasRise2020 provided information for the number of homeless individuals during the COVID 19 pandemic a more current report can be found at https://www.statista.com/statistics/727847/homelessness-rate-in-the-us-by-state/]. With the current Covid-19 pandemic, we can only predict that those numbers have continued to increase. In the United States, 2.4% of homeless individuals die each year (Stasha, 2020). We know that the general population often tries to distance itself from the stigmatized population, more specifically the homeless population. Homeless individuals face greater stigma and social isolation and often are removed from public parks and other locations because the general public does not want them too close. The problems caused by stigmatization, such as social distancing, can affect the homeless population in terms of resources that they have available such as sanitation centers, employment, and social support. Often the homeless population lacks resources and is exposed to the elements which can increase their mortality, as well as the chance of being malnourished, having parasitic infestations, periodontal disease, degenerative joint diseases, venereal diseases, cirrhosis, and hepatitis-related to intravenous (IV) drug abuse. Public attitudes toward homeless individuals can influence policies and the services provided to this population. The attitudes displayed through the stigma of homeless individuals can have an impact on both physical and psychological health and willingness to access services. The impact of these stigmas has shown to have serious implications on the well-being of homeless individuals. The present study examined factors that could predict levels of stigmatization expressed towards homeless individuals. 

## Literature review 

Research on the stigma of mental illness, homelessness, and race highlights its harmful effects on health and social integration. @corriganPublicStigmaMental2009 examined public stigma, focusing on stereotypes like causal attribution (blaming individuals for their condition) and dangerousness (perceiving them as threatening). Using vignette-based experiments, the study found that people with psychiatric disorders, especially those with drug addiction, faced greater stigma than those with physical disabilities, laying the foundation for understanding how schizophrenia and substance use disorders contribute to homelessness stigma.

While @corriganAttributionModelPublic2003 explored mental illness stigma, it did not examine health outcomes. In contrast, @weiszStigmatizedIdentitiesPsychological2018 demonstrated that homelessness stigma leads to psychological distress, poor health, and social avoidance. Among 175 volunteers at a homeless event, those experiencing or anticipating stigma reported worse physical and mental health and greater reluctance to seek services. Participants of color faced even higher distress and service avoidance, highlighting the compounded impact of race and homelessness stigma.

Building on this, @markowitzRaceGenderHomelessness2021 investigated race and gender intersections in stigma. They found that black homeless individuals were perceived as more dangerous than white counterparts, though no significant differences in social distance emerged. However, the study's reliance on college-aged participants, who may have been more tolerant than the general population, was a limitation. The present study addresses this by including a broader, more diverse sample.

Similarly, @gattisPerceivedRacialSexual2016 linked racial stigma and discrimination to heightened depression among 89 black adolescents and young adults experiencing homelessness. Using social and minority stress models, the study highlighted how marginalized groups face greater psychological distress due to limited societal support. Though constrained by a small sample, it reinforced the role of racial stigma in homelessness experiences.

In sum, stigma related to homelessness, mental illness, and race profoundly affects psychological and physical health, social integration, and resource access. The present study expands on this research by examining how mental illness and race interact to shape stigma, offering a more nuanced understanding of its impact on homeless individuals 

## Current Study
Building on the research by @corriganPublicStigmaMental2009, @markowitzRaceGenderHomelessness2021, and @weiszStigmatizedIdentitiesPsychological2018 the current study aimed to explore how race and mental illness diagnosis impact the stigmatization of homeless individuals. The research specifically focused on mental illness, distinguishing between individuals with schizophrenia and those with substance use disorders, and examined how these factors interact with race in shaping stigma. Previous studies suggest that public stigma varies across mental health conditions and that race plays a crucial role in determining the intensity of stigma. Based on these findings, the present study hypothesized that race would significantly influence social distance, perceived danger, blameworthiness, and emotional responses (concern and help) toward homeless individuals. Specifically, it was predicted that black homeless individuals would experience greater social distance, be perceived as more dangerous and more blameworthy, and receive less concern and help compared to their white counterparts. Additionally, it was anticipated that individuals with substance use disorders would face higher levels of social distance, dangerousness, and blame, while individuals with schizophrenia would receive more concern and help. Lastly, the study predicted that race and mental illness diagnosis would interact to influence all aspects of stigmatization.

### Hypotheses 
#### Effect of Race on Stigmatization
> 1. Black homeless individuals will experience greater social distance.
> 2. Black homeless individuals will be perceived as more dangerous.
> 3. Black homeless individuals will be perceived as more blameworthy.
> 4. Black homeless individuals will receive less concern and help compared to white homeless individuals.

#### Effect of Mental Illness Diagnosis on Stigmatization

> 1. Individuals with substance use disorders will face higher levels of social distance.
> 2. Individuals with substance use disorders will be perceived as more dangerous.
> 3. Individuals with substance use disorders will be perceived as more blameworthy.
> 4. Individuals with schizophrenia will receive more concern and help.
> 5. Interaction Between Race and Mental Illness Diagnosis

#### Race and mental illness diagnosis interaction
> 1. Race and mental Illness diagnosis will interact to influence all aspects of stigmatization, including social distance, perceived danger, blameworthiness, concern, and willingness to help.

# Method

## Participants

The sample for this study consisted of 215 participants, primarily college-aged students in the United States, with ages ranging from The ages of the sample ranged from 18 to 79 (M = `r round(35.08, 2)`, SD = `r round(16.33, 2)`).The sample was 44.7% White, 13% Hispanic, 33% Black, 4.2% Asian American, 3.3% Biracial, and 1.9% other ethnicities. The sample identified politically as 52.6% Liberal, 37.1% Moderate, and 10.3% Conservative.The sample was broken down into 1.9% living in a rural community, 58.6% living in the suburbs, 12.1% living in a small town, and 27.4% living in a large metropolitan city. Finally, the distribution of gender was as follows: 29.3% male, 68.8% female, and 1.9% other responses. **Table 1.** provides a summary of the demographic characteristics of the sample.

### Table 1
Demographic Information of Participants

| **Demographic**                    | **Percentage** |
|------------------------------------|----------------|
| **Age (M = 35.08, SD = 16.33)**    | -              |
| **Ethnicity**                      |                |
| White                              | 44.7%          |
| Hispanic                           | 13%            |
| Black                              | 33%            |
| Asian American                     | 4.2%           |
| Biracial                           | 3.3%           |
| Other Ethnicities                  | 1.9%           |
| **Political Affiliation**          |                |
| Liberal                            | 52.6%          |
| Moderate                           | 37.1%          |
| Conservative                       | 10.3%          |
| **Location**                       |                |
| Rural Community                    | 1.9%           |
| Suburbs                            | 58.6%          |
| Small Town                         | 12.1%          |
| Large Metropolitan City            | 27.4%          |
| **Gender**                         |                |
| Male                               | 29.3%          |
| Female                             | 68.8%          |
| Other Responses                    | 1.9%           |

<p style="text-align: center; font-style: italic;">**Note**: Percentages may not sum to 100 due to rounding.</p>

## Measures
This study used multiple questionnaires to examine the effects and interactions of race and mental illness on stigmatization toward homeless individuals. Participants were assigned to one of six experimental conditions using vignettes adapted from @markowitzRaceGenderHomelessness2021, manipulating race and mental illness. **Table 2.** presents the vignettes used in this study. 

### Table 2
Vignettes Used in the Study

| **Condition**                                        | **Race**   | **Mental Illness**       | **Character Description**                             |
|------------------------------------------------------|------------|--------------------------|-------------------------------------------------------|
| Condition 1: Black character/No mental illness       | Black      | No mental illness         | Male homeless individual with same life story         |
| Condition 2: Black character/Substance use disorder  | Black      | Substance use disorder    | Male homeless individual with same life story         |
| Condition 3: Black character/Schizophrenia           | Black      | Schizophrenia             | Male homeless individual with same life story         |
| Condition 4: White character/No mental illness      | White      | No mental illness         | Male homeless individual with same life story         |
| Condition 5: White character/Substance use disorder  | White      | Substance use disorder    | Male homeless individual with same life story         |
| Condition 6: White character/Schizophrenia          | White      | Schizophrenia             | Male homeless individual with same life story         |

**Note:** All vignettes used the same life story for the male homeless individual, with only race and mental illness varying across conditions.

### Attribution Questionnaire
 The Attribution Questionnaire[@corriganAttributionModelPublic2003] assessed stigmatization aspects like social distance, blame, perceived dangerousness, emotional response, and willingness to help. 

### Memory Check
A Memory Check assessed participants' recall of story details, specifically the race and mental illness of the character. 

### Demographic Questionnaire
Demographics questionnaire that asked participants about their age, ethnicity, residence, political affiliation, and familiarity with homelessness.

# Results
## Descriptive Statistics
```{r,echo=TRUE,include=FALSE}
# Load necessary libraries
library(readr)

# Read the CSV file
data <- read_csv("Data_Stigmatization_of HomelessIndividuals/Experimental_Study_Karen.csv")

# View the structure of the data
str(data)
```

```{r,echo=TRUE,include=FALSE}
# Summary statistics for age
summary(data$age)

# Summary statistics for Distance_mean
summary(data$Distance_mean)

# Summary statistics for blame_mean
summary(data$Blamemean)

# Summary statistics for danger_mean
summary(data$danger_mean)

# Summary statistics for concern_mean
summary(data$Concern_mean)

# Summary statistics for help_mean
summary(data$Help_Mean)

```

```{r,echo=TRUE,include=FALSE}
# Frequency and proportions for gender
table(data$gender)
prop.table(table(data$gender))

# Frequency and proportions for ethnicity
table(data$ethnicity)
prop.table(table(data$ethnicity))

# Frequency and proportions for political_affiliation
table(data$political_affiliation)
prop.table(table(data$political_affiliation))

# Frequency and proportions for community
table(data$community)
prop.table(table(data$community))

```

```{r,echo=TRUE,include=FALSE}
# Summary, mean, median, and frequency for socialdistance_1
summary(data$socialdistance_1)
mean(data$socialdistance_1, na.rm = TRUE)
median(data$socialdistance_1, na.rm = TRUE)
table(data$socialdistance_1)

# Summary, mean, median, and frequency for blame_1
summary(data$blame_1)
mean(data$blame_1, na.rm = TRUE)
median(data$blame_1, na.rm = TRUE)
table(data$blame_1)

# Summary, mean, median, and frequency for danger_1
summary(data$danger_1)
mean(data$danger_1, na.rm = TRUE)
median(data$danger_1, na.rm = TRUE)
table(data$danger_1)

# Summary, mean, median, and frequency for concern_1
summary(data$concern_1)
mean(data$concern_1, na.rm = TRUE)
median(data$concern_1, na.rm = TRUE)
table(data$concern_1)

# Summary, mean, median, and frequency for help_1
summary(data$help_1)
mean(data$help_1, na.rm = TRUE)
median(data$help_1, na.rm = TRUE)
table(data$help_1)

```

```{r,echo=TRUE,include=FALSE}
# Summary statistics results for age
summary_age <- summary(data$age)
age_min <- round(summary_age[1], 2)
age_1st_quartile <- round(summary_age[2], 2)
age_median <- round(summary_age[3], 2)
age_mean <- round(summary_age[4], 2)
age_3rd_quartile <- round(summary_age[5], 2)
age_max <- round(summary_age[6], 2)

# Summary statistics results for Distance_mean
summary_distance <- summary(data$Distance_mean)
distance_min <- round(summary_distance[1], 2)
distance_1st_quartile <- round(summary_distance[2], 2)
distance_median <- round(summary_distance[3], 2)
distance_mean <- round(summary_distance[4], 2)
distance_3rd_quartile <- round(summary_distance[5], 2)
distance_max <- round(summary_distance[6], 2)

# Summary statistics results for blame_mean
summary_blame <- summary(data$Blamemean)
blame_min <- round(summary_blame[1], 2)
blame_1st_quartile <- round(summary_blame[2], 2)
blame_median <- round(summary_blame[3], 2)
blame_mean <- round(summary_blame[4], 2)
blame_3rd_quartile <- round(summary_blame[5], 2)
blame_max <- round(summary_blame[6], 2)

# Summary statistics results for danger_mean
summary_danger <- summary(data$danger_mean)
danger_min <- round(summary_danger[1], 2)
danger_1st_quartile <- round(summary_danger[2], 2)
danger_median <- round(summary_danger[3], 2)
danger_mean <- round(summary_danger[4], 2)
danger_3rd_quartile <- round(summary_danger[5], 2)
danger_max <- round(summary_danger[6], 2)

# Summary statistics results for concern_mean
summary_concern <- summary(data$Concern_mean)
concern_min <- round(summary_concern[1], 2)
concern_1st_quartile <- round(summary_concern[2], 2)
concern_median <- round(summary_concern[3], 2)
concern_mean <- round(summary_concern[4], 2)
concern_3rd_quartile <- round(summary_concern[5], 2)
concern_max <- round(summary_concern[6], 2)

# Summary statistics results for help_mean
summary_help <- summary(data$Help_Mean)
help_min <- round(summary_help[1], 2)
help_1st_quartile <- round(summary_help[2], 2)
help_median <- round(summary_help[3], 2)
help_mean <- round(summary_help[4], 2)
help_3rd_quartile <- round(summary_help[5], 2)
help_max <- round(summary_help[6], 2)

# Results text for manuscript this was super fun to make. I would like to do this more
cat("
The summary statistics for the age variable showed a minimum value of", age_min, ", first quartile of", age_1st_quartile, ", median of", age_median, ", mean of", age_mean, ", third quartile of", age_3rd_quartile, ", and a maximum of", age_max, ".\n")

cat("
For Distance_mean, the minimum value was", distance_min, ", first quartile was", distance_1st_quartile, ", median was", distance_median, ", mean was", distance_mean, ", third quartile was", distance_3rd_quartile, ", and the maximum was", distance_max, ".\n")

cat("
The Blamemean variable had a minimum value of", blame_min, ", first quartile of", blame_1st_quartile, ", median of", blame_median, ", mean of", blame_mean, ", third quartile of", blame_3rd_quartile, ", and a maximum of", blame_max, ".\n")

cat("
For danger_mean, the minimum was", danger_min, ", first quartile was", danger_1st_quartile, ", median was", danger_median, ", mean was", danger_mean, ", third quartile was", danger_3rd_quartile, ", and the maximum was", danger_max, ".\n")

cat("
The concern_mean had a minimum of", concern_min, ", first quartile of", concern_1st_quartile, ", median of", concern_median, ", mean of", concern_mean, ", third quartile of", concern_3rd_quartile, ", and a maximum of", concern_max, ".\n")

cat("
Finally, for help_mean, the minimum was", help_min, ", first quartile was", help_1st_quartile, ", median was", help_median, ", mean was", help_mean, ", third quartile was", help_3rd_quartile, ", and the maximum was", help_max, ".\n")

```
We ran some summary statistics for the sample focusing on it was inprtant for us to highlight that our sample was composed of mostly a college age sample despite being open to anyone that wanted to participate. We highlight this as a limiation to our work.For the age showed a minimum value of `r age_min`, a first quartile of `r age_1st_quartile`, a median of `r age_median`, a mean of `r age_mean`, a third quartile of `r age_3rd_quartile`, and a maximum of `r age_max`.

In addtion we ran summary statistics for the variable of social dictance that will be focused in this analysis.For Distance_mean , the minimum value was `r distance_min`, the first quartile was `r distance_1st_quartile`, the median was `r distance_median`, the mean was `r distance_mean`, the third quartile was `r distance_3rd_quartile`, and the maximum was `r distance_max`.The results for this variable are presented in the following section in which an ANOVA was conduected to examine if there were any significant main effects or interactions between race and mental illness diagnosis on social distance stigma.

## Social Distance Stigma
The ANOVA results show that the main effect of diagnosis on social distance stigma showed a significant effect of diagnosis on social distance stigma, F(1, 211) = `r round(40.31, 2)`, p < .001. However, there was no significant interaction between race and diagnosis on social distance stigma, F(1, 211) = `r round(0.87, 2)`, p = `r round(0.512, 3)`. Similarly, the main effect of race on social distance stigma did not reach significance, F(1, 211) = `r round(0.07, 2)`, p = `r round(0.850, 3)`. **@fig-anova-result** shows the F-values for the main effects and interaction.

```{r}
#| label: anova-analysis
#| echo: true
#| results: hide
#| include: false

library(tidyverse)

# Read the dataset
Experimental_Study_Karen <- read_csv("Data_Stigmatization_of HomelessIndividuals/Experimental_Study_Karen.csv")

# Conducting the 2x3 between-subjects ANOVA
anova_results <- aov(socialdistance_1 ~ IV1_race * IV2_diagnosis, data = Experimental_Study_Karen)

# Display ANOVA summary
summary(anova_results)

```

```{r}
#| label: fig-anova-result
#| fig-cap: ANOVA F-values for Race, Diagnosis, and Their Interaction.
#| apa-note: The plot shows F-values for each predictor in the ANOVA model. Higher F-values indicate a stronger effect on the dependent variable.
#| #| echo: FALSE
#| fig.pos: "H"
#| fig-width: 7
#| fig-height: 5

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tibble)

# Data for plotting
anova_results <- tibble(
  Predictor = c("Race (IV1)", "Diagnosis (IV2)", "Race × Diagnosis"),
  F_Value = c(0.036, 19.930, 0.431),
  p_Value = c(0.850, 1.31e-05, 0.512)
)

# Add significance labels
anova_results <- anova_results %>%
  mutate(Significance = case_when(
    p_Value < 0.001 ~ "***",
    p_Value < 0.01 ~ "**",
    p_Value < 0.05 ~ "*",
    TRUE ~ ""
  ))

# Create bar plot
ggplot(anova_results, aes(x = Predictor, y = F_Value, fill = Predictor)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_text(aes(label = Significance), vjust = -0.5, size = 5) +  # Add significance labels
  labs(
    x = "Predictor",
    y = "F-Value",
    title = "ANOVA Results for Race, Diagnosis, and Their Interaction"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```
### Mean Stigma Scores by Diagnosis
```{r}
#| label: fig-social-distance-stigma
#| fig-cap: Effect of Mental Illness Diagnosis on Social Distance Stigma.
#| apa-note: The plot shows mean social distance stigma scores by diagnosis, with error bars representing standard deviation.
#| #| echo: FALSE
#| fig.pos: "H"
#| #| fig-width: 7
#| fig-height: 5

# data for plotting
data <- tibble(
  diagnosis = c("No Diagnosis", "Schizophrenia", "Substance Use Disorder"),
  mean_stigma = c(2.58, 3.40, 3.24),
  sd_stigma = c(1.12, 1.21, 1.26)
)

# Create bar plot
ggplot(data, aes(x = diagnosis, y = mean_stigma, fill = diagnosis)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean_stigma - sd_stigma, ymax = mean_stigma + sd_stigma), width = 0.2) +
  labs(x = "Mental Illness Diagnosis", y = "Mean Social Distance Stigma",
       title = "Impact of Mental Illness Diagnosis on Social Distance Stigma") +
  theme_minimal(base_size = 10)
```
**@fig-social-distance-stigma** shows the effect of mental illness diagnosis on social distance stigma. The mean stigma scores for each diagnosis are as follows:No Diagnosis: `r data$mean_stigma[1]` (SD = `r data$sd_stigma[1]`), Schizophrenia: `r data$mean_stigma[2]` (SD = `r data$sd_stigma[2]`),Substance Use Disorder: `r data$mean_stigma[3]` (SD = `r data$sd_stigma[3]`)

The results indicate that individuals with schizophrenia (M = `r data$mean_stigma[2]`) and substance use disorder (M = `r data$mean_stigma[3]`) experience higher levels of social distance stigma compared to those with no diagnosis (M = `r data$mean_stigma[1]`). These differences were statistically significant.

### Correlations
```{r,echo=TRUE,include=FALSE}
#| label: correlation-analysis
# Load necessary library
library(dplyr)
#Read the dataset
df <- read.csv("Data_Stigmatization_of HomelessIndividuals/cleaned_stigma_data.csv") # using the cleaned data set 

# Perform Pearson correlation test
cor.test(df$Social_Desirability_Bias, df$Mean_Help_Offering, method = "pearson")# both are numeric data
```

```{r}
#| label: fig-correlation-analysis-help-and-social-desirability
#| fig-cap: Correlation between social desirability bias and mean help offering
#| apa-note: The plot shows the relationship between social desirability bias and mean help offering, with a regression line indicating the trend.
#| echo: FALSE
#| results: "hide"
#| fig.pos: "H"
#| #| fig-width: 7
#| fig-height: 5

# Load necessary libraries
library(ggplot2)

# Load the data
df <- read.csv("Data_Stigmatization_of HomelessIndividuals/cleaned_stigma_data.csv")

# Fit the linear model
model <- lm(Mean_Help_Offering ~ Social_Desirability_Bias, data = df)

# color gradient and direction of correlation line
plot <- ggplot(df, aes(x = Social_Desirability_Bias, y = Mean_Help_Offering, color = Social_Desirability_Bias)) +
  geom_point(size = 3) +  #  point size
  # Conditionally remove geom_smooth based on whether it's being rendered
  {if (knitr::is_html_output()) geom_smooth(method = "lm", se = FALSE, color = "red")} +  # Red regression line 
  geom_abline(intercept = coef(model)[1], slope = coef(model)[2], color = "blue", linewidth = 1) +  # Line showing correlation direction
  scale_color_gradient(low = "lightblue", high = "darkblue") +  # Gradient from light blue to dark blue
  labs(
    title = "Social Desirability Bias and Mean Help Offering",
    x = "Social Desirability Bias",
    y = "Mean Help Offering"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 9)
  )

# Print plot
print(plot)

```
The correlational analysis conducted in this section are to guide where future research should focus given that there could be other factors that are influencing the stigmatization of homeless individuals.

A Pearson correlation test was conducted to examine the relationship between social desirability bias and mean help offering. The results showed a significant negative correlation between social desirability bias and mean help offering, r = `r round(-0.28, 2)`, p < .001, indicating that participants with higher social desirability bias were less likely to offer help to homeless individuals.**@fig-correlation-analysis-help-and-social-desirability** shows the relationship between social desirability bias and mean help offering, with a regression line indicating the trend.
```{r}
#| label: fig-correlation-age-and-mean-help
#| fig-cap: Correlation between age of participants and mean help offering
#| apa-note: The plot shows the relationship between age of participants and mean help offering, with a regression line in blue indicating the trend.The red line is the mean. The correlation coefficient quantifies the strength and direction of this relationship.
#| echo: FALSE
#| results: "hide"
#| fig.pos: "H"
#| fig-width: 7
#| fig-height: 5

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the data
df <- read.csv("Data_Stigmatization_of HomelessIndividuals/cleaned_stigma_data.csv")

# Clean the data (handle NAs)
df_clean <- df %>%
  drop_na(Age_of_Participant, Mean_Help_Offering)  # Ensure no NAs in important variables

# Calculate the correlation between Age_of_Participant and Mean_Help_Offering
correlation_result <- cor(df_clean$Age_of_Participant, df_clean$Mean_Help_Offering, method = "pearson")

# Print the Pearson correlation coefficient
print(paste("Pearson correlation coefficient: ", round(correlation_result, 2)))

# Detailed summary with significance testing
cor_test_result <- cor.test(df_clean$Age_of_Participant, df_clean$Mean_Help_Offering)

# Print detailed results
print(cor_test_result)

# Fit a linear model
model <- lm(Mean_Help_Offering ~ Age_of_Participant, data = df_clean)

# aesthetics
plot <- ggplot(df_clean, aes(x = Age_of_Participant, y = Mean_Help_Offering)) +
  geom_point(size = 3) +  # Point size
  # Conditionally add geom_smooth for HTML output
  {if (knitr::is_html_output()) geom_smooth(method = "lm", se = FALSE, color = "red")} +  # Red regression line 
  geom_abline(intercept = coef(model)[1], slope = coef(model)[2], color = "blue", linewidth = 1) +  # Line showing correlation direction
  geom_hline(yintercept = mean(df_clean$Mean_Help_Offering, na.rm = TRUE), 
             linetype = "solid", color = "red", linewidth = 1.2) +  # Add horizontal line for mean
  labs(
    title = "Age of Participants and Mean Help Offering",
    x = "Age of Participant",
    y = "Mean Help Offering"
  ) +
  theme_minimal() +  # Keep minimal theme
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10)
  )

# Print the plot
print(plot)
```

```{r,echo=TRUE,include=FALSE}
# Load necessary libraries
library(dplyr)

# Load the data
df <- read.csv("Data_Stigmatization_of HomelessIndividuals/cleaned_stigma_data.csv")

# Clean the data (handle NAs)
df_clean <- df %>%
  drop_na(Age_of_Participant, Mean_Help_Offering)  # Ensure no NAs in important variables

# Perform the correlation test
cor_test_result <- cor.test(df_clean$Age_of_Participant, df_clean$Mean_Help_Offering)

# Print the p-value
cor_test_result$p.value

```
Finally, we conducted a correlation analysis to examine the relationship between participants' age and their mean help offering. The Pearson correlation coefficient between age and mean help offering was `r round(correlation_result, 2)`, quantifying the strength and direction of this relationship in this case a negative relationship.

The results indicated a significant correlation, r = `r round(correlation_result, 2)`, *p* = `r signif(cor_test_result$p.value, 3)`, suggesting that age is meaningfully associated with help offering.
**@fig-correlation-age-and-mean-help** shows the relationship between participants' age and their mean help offering, with a regression line indicating the trend. The red line represents the mean help offering, while the blue line shows the direction of the correlation.


# Discussion
This study explored how race and mental illness diagnosis impact the stigmatization of homeless individuals, specifically focusing on social distance. The results revealed that race did not significantly affect social distance stigma, suggesting that race was not a key factor in determining social distance in this context. However, the diagnosis of mental illness had a significant impact on social distance. Individuals with schizophrenia and substance use disorder were perceived as more socially distant than those with no diagnosis, supporting the hypothesis that mental illness contributes to higher levels of social distance stigma.

The absence of a significant interaction between race and diagnosis further highlights that mental illness plays a larger role in shaping social distance perceptions than race. This finding suggests that, when it comes to social distance, mental illness may be a more prominent factor than race in the stigmatization of homeless individuals.

These findings underscore the importance of addressing mental illness as a key determinant in social stigma, particularly in the context of homelessness. Further research is needed to explore how these factors interact in other forms of stigmatization and to investigate the role of mental illness across different populations.

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
